GLM Predictor Matrices

### 2020-12-12 workflow/scripts/predictor_matrices.Rmd
wip predictor matrices

```{r, echo = FALSE, results ='hide', message = FALSE}
library(knitr)
library(wpp2019)
library(tidyverse)
library(readr)
source("./scripts/flight_data.R")
```

@@

```{r}
## Daily flight passengers

demes <- c(China = "CN", France = "FR", Germany = "DE", Italy = "IT", OtherEuropean = "OE", Spain = "ES")

@@

fligh_data_3i <- fligh_data %>%
  filter((MONTH %in% 1:3 & YEAR == 2020) | (MONTH == 12 & YEAR == 2019)) %>%
  mutate(EPOCH = case_when(
    MONTH == 2 ~ 2,
    MONTH == 3 ~ 3,
    TRUE ~ 1)) %>%
  group_by(GEO, PARTNER, EPOCH) %>%
  mutate(N = n(),
         EVALUE = sum(DVALUE/N)) %>%
  ungroup %>%
  # Transformations
  mutate(LVALUE = log(EVALUE + 1),
         SLVALUE = scale(LVALUE))

write(slv_flightMatrix_3i, "files/flight_matrix_3i.csv", sep = ",")
```

@@

Population counts from World Population Prospects 2019, United Nations Population Division,  `wpp2019` R library. Values for **China**, **France**, **Germany**, **Italy**, **Other European** and **Spain**.

@@

\small
```{r, echo=FALSE}
data(pop)
population <- pop %>% filter(name %in% labels(demes)) %>% select(name, "2020") 
pop_eu_demes <- population %>% filter(name != "China") %>% summarise(s = sum(`2020`)) %>% pull(s)
pop_eu_total <- pop %>% filter(name == "Europe") %>% pull("2020")
population <- population %>%
  tibble::add_row(name = "OtherEuropean", `2020` = pop_eu_total - pop_eu_demes) %>%
  arrange(name)
```

```{r, echo=FALSE}
# Population source matrix
pop_source <- matrix(1, 6, 6) * population$`2020`
diag(pop_source) <- NA
vslpop_source <- scale(log(c(pop_source)))
slpop_source <- matrix(vslpop_source, 6, 6)
colnames(slpop_source) <- demes[population$name]
rownames(slpop_source) <- demes[population$name]

@@

## Distance between countries

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

>>>>

```{r, echo = FALSE, results ='hide', message = FALSE}
library(knitr)
library(wpp2019)
library(tidyverse)
library(readr)
library(ggsci)
library(scales)
source("./scripts/utils.R")
source("./scripts/flight_data.R")
```

@@

## GLM Model

We define the migration rates $m_{ij}^{e}$ from deme $i$ to deme $j$ in epoch $e$ as a log linear function of a set of predictors $\boldsymbol{X}$, log transformed and scaled, such that

$$
m_{ij}^{e} = c \exp(\beta_{pas} \delta_{pas} x_{ij.pas}^e + \beta_{pop} \delta_{pop }x_{ij.pop} + \beta_{dist} \delta_{dist}x_{ij.dist})
$$

where $\boldsymbol{\beta}$ are the coefficients for the predictors that can be between $-\infty$ and $\infty$ (prior  $\mathcal{N}(0,4)$) and $\boldsymbol{\delta}$ are the indicators that can be 0 or 1 and denote if a predictor contributes at all.

## Daily flight passengers

Passengers carried data from Eurostats. Reported _departures_ values are taken for European countries (recommendation from Eurostats to avoid duplicates). For China, arrivals and departures values are considered.

- **Epoch 1:** December 2019 - January 2020 (Wuhan lockdown 23 Jan)
- **Epoch 2:** February 2020 
- **Epoch 3:** March 2020

\tiny
```{r, echo = FALSE, message = FALSE, warning = FALSE}
demes <- c(China = "CN", France = "FR", Germany = "DE", Italy = "IT", OtherEuropean = "OE", Spain = "ES")

@@

fligh_data_3i <- fligh_data %>%
  filter((MONTH %in% 1:3 & YEAR == 2020) | (MONTH == 12 & YEAR == 2019)) %>%
  mutate(EPOCH = case_when(
    MONTH == 2 ~ 2,
    MONTH == 3 ~ 3,
    TRUE ~ 1)) %>%
  group_by(GEO, PARTNER, EPOCH) %>%
  summarise(N = n(),
         EVALUE = sum(DVALUE/N)) %>%
  ungroup %>%
  # Transformations
  mutate(LVALUE = log(EVALUE + 1),
         SLVALUE = scale(LVALUE)) %>%
  arrange(EPOCH, factor(GEO, levels = demes), factor(PARTNER, levels = demes))

#write(slv_flightMatrix_3i, "files/flight_matrix_3i.csv", sep = ",")
```

```{r, echo = FALSE}
flight_matrices <- lapply(1:3, function (i) {
  fligh_data_3i %>%
  filter(EPOCH == i) %>%
  select(GEO, PARTNER, SLVALUE) %>%
  pivot_wider(names_from = "PARTNER", values_from = "SLVALUE") %>%
  mutate_if(is.numeric, ~round(.,2)) %>%
  column_to_rownames("GEO") %>%
  select(unname(demes))
})

knitr::kable(flight_matrices, booktabs = TRUE, caption = "Flight passengers between demes, epochs 1-3")
```

## Daily flight passengers

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 3}
set_plotopts()
dcolors <- pal_npg("nrc")(10)[c(1:5,9)]
dcolors <- dcolors[1:length(demes)]
names(dcolors) <- demes[sort(labels(demes))]
#dark_dcolors <- rgb(t(col2rgb(dcolors)/2), maxColorValue=255)
#names(dark_dcolors) <- names(dcolors)

# Fligh data plot
ggplot(fligh_data_3i, aes(x = EPOCH, y = EVALUE, color = PARTNER)) +
  geom_line() +
  geom_point(size = 1) +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x)), breaks = trans_breaks('log10', function(x) 10^x)) +
  scale_x_continuous(breaks = c(1, 2, 3), labels = c("E1", "E2", "E3")) +
  facet_grid(~GEO) +
  xlab("Epoch") +
  ylab("# Daily Flight Passengers") + 
  #theme(axis.title.x = element_blank()) + 
  scale_color_manual(values = dcolors, name = "Destination")
```

@@

```{r, echo=FALSE}
data(pop)
population <- pop %>% filter(name %in% labels(demes)) %>% select(name, "2020") 
pop_eu_demes <- population %>% filter(name != "China") %>% summarise(s = sum(`2020`)) %>% pull(s)
pop_eu_total <- pop %>% filter(name == "Europe") %>% pull("2020")
population <- population %>%
  tibble::add_row(name = "OtherEuropean", `2020` = pop_eu_total - pop_eu_demes) %>%
  arrange(name)
```

Population counts from World Population Prospects 2019, United Nations Population Division,  `wpp2019` R library: **China** - `r population[1,2] * 1e3`, **France** - `r population[2,2] * 1e3`, **Germany** - `r population[3,2] * 1e3`, **Italy** - `r population[4,2] * 1e3`, **Other European** - `r population[5,2] * 1e3` and **Spain** - `r population[6,2] * 1e3`.

@@

\tiny
```{r, echo=FALSE}
# TODO use long format, easy to combine later to export to xml, like with the flight data
# Population source matrix
pop_source <- matrix(1, 6, 6) * population$`2020`
diag(pop_source) <- NA
vslpop_source <- scale(log(c(pop_source)))
slpop_source <- matrix(vslpop_source, 6, 6)
colnames(slpop_source) <- demes[population$name]
rownames(slpop_source) <- demes[population$name]

@@

## Distance between countries
Options:
Great circle distance between centroids of each country - Which distance for other european deme? Could be the average
Average pairwaise distance between airports - considers that only flight transportation matters, more work
Distance taking into account population distribution inside the country.
Minimum distance.
